
---
title: "Assignment 5: Medicaid Expansion and the Difference-in-differences Design"
subtitle: "Part 1: Single Treatment Timing - SOLUTIONS"
author: "Instructor: Harold Stolper"  
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 1
    number_sections: true
    code_folding: hide
    df_print: paged
header-includes:
- \usepackage{amsmath}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


<style type="text/css">

h2 { /* Header 2 */
    font-size: 18px;
    font-weight: bold;
}
</style>

<br>

This lesson closely follows [Difference-in-Differences Designs: A Practitioner’s Guide](https://arxiv.org/abs/2503.13323) by Baker et al (2025).

___

#### <span style="text-decoration:underline">Instructions</span>: Write up your answers in the provided rmd file, make sure to maintain the same numbering in this document to ensure proper credit for your responses. Your submission should include: ***(1) all R code and output you used to answer each question in your document;*** and **(2) written answers to each question along with your reasoning.** {.unnumbered}

#### Unlike previous assignments with **pdf** submissions, for this one you will **submit an html file** knit from an rmd file, along with the rmd file. {.unnumbered}

#### **This means you can work directly from this rmd file rather than from a separate R script**. After installing any necessary packages, start by test knitting the rmd file with the assignment questions as an html file. As you work through the assignment, modify/add code chunks as necessary and add your write up. Remember to test knit frequently as you add code. To more easily explore the data without knitting, you will also need to run all of the code such that the output appears in your current environment in the upper right hand panel (and not just your knitting environment). To do this, click the green play button at the top of every code chunk as you go. {-}

#### You will also need to install several any new packages listed below in order to knit this document, namely `did`, `HonestDID` and `latex2exp`. {.unnumbered}

```{r}
library(tidyverse)
library(kableExtra)
library(modelsummary)
library(did)
library(fixest)
library(HonestDiD)
library(latex2exp)
#library(gt)
#library(ggthemes)
```
___
<br>

# A Note on Research Design {.unnumbered}

Before we proceed, let's talk briefly about research design. Think back to Quant II, where we focused on using Ordinary Least Squares to generate estimates of a parameter from a population regression function (PRF).  We wrestled with the conditions that needed to hold for the parameter estimate to represent an unbiased effect of the explanatory variable of interest on the outcome variable. 

Now let's think more carefully about the target parameter we would like to estimate. We can actually think about different causal parameters, sometimes referred to as *estimands*; for example, the average treatment effect for the entire population (ATE), or, more narrowly, the average treatment effect on the treated (ATT). Of course for any individual entity, we never observe the outcome under both a treated and untreated state. This is the fundamental challenge with causal inference: we never observe the counterfactual! So we need a plan to mimic the counterfactual for some individual entities and arrive at an unbiased estimate of a causal parameter.

This is what we are doing with research design: research design is really just a strategy and set of assumptions to identify and estimate a target parameter for a given set of data. This lesson will help us understand a specific research design that is quite common: identifying and estimating the ATT of a binary treatment that has a staggered rollout across entities—the staggered difference-in-difference (DID) design. However, Part 1 will develop the framework under a simpler DID design with a single treatment date for all treated units, not a staggered rollout.


# Introduction

In this lesson we will explore the same causal question with same data as Baker et al (2025): How did the expansion of public health insurance in the United States (Medicaid) under the Affordable Care Act (ACA) affect mortality? 

Medicaid is a long-running federal and state program that provides free or low-cost healthcare to Americans with limited income and resources. The ACA, signed in 2010, is a health care reform law designed to make healthcare more affordable. One provision of the ACA enabled states, starting in 2014, to raise the Medicaid income eligibility threshold and reach relatively higher income households.   

Medicaid expansion across the states showcases a staggered treatment adoption: despite a mandate by the ACA in 2014 that all states expand Medicaid eligibility to adults with incomes up to 138% of the federal poverty threshold, the Supreme Court made Medicaid expansion optional. Many states chose to expand Medicaid after 2014, but several states still did not so as of 2024. States chose to expand or not expand Medicaid due to a variety of economic and political considerations (Sommers and Epstein, 2013), resulting in observable differences between expansion and non-expansion states. 


```{r}
# load cleaned data
load('bakeretal_cleaned.rdata')

# set seed for reproducibility
set.seed(20240924)
```


## QUESTION: Using the `adopt` variable, report the adoption year by state.

::: {style="color:deepskyblue;"}
**Solution:**
```{r}
# get adoption year by state
mydata %>% 
  select(state, adopt) %>% 
  distinct() %>% 
  arrange(state)
```
:::

<br>

Variation in medication expansion dates is summarized in the following table:
```{r}
# first get the share of states, share of counties, and share of adults in 2013 by adoption category
# first the states and share of states
states <- mydata %>% 
  select(state, adopt) %>% 
  distinct() %>% 
  arrange(state) %>% 
  group_by(adopt) %>% 
  summarize(states = paste0(state, collapse = ", "),
            state_share = length(state) / 50)

# next get the county share and the population share
counties_pop <- mydata %>% 
  # just for year 2013
  filter(year == 2013) %>% 
  # get county and population totals
  mutate(total_counties = n(),
         total_pop = sum(population_20_64, na.rm = TRUE)) %>% 
  group_by(adopt) %>% 
  # make into shares
  summarize(county_share = n() / mean(total_counties),
            pop_share = sum(population_20_64, na.rm = TRUE) / mean(total_pop))

# make a table and export
states %>% 
  left_join(counties_pop, by = "adopt") %>% 
  slice(9, 1:8) %>% 
  # format the numbers to two digits
  mutate(across(state_share:pop_share, ~ scales::number(., accuracy = 0.01))) %>%
  # format the table
  kable(
    col.names = c("Expansion Year", "States", "Share of States", "Share of Counties", 
                  "Share of Adults (2013)"),
    booktabs = T, caption = "Medicaid Expansion Under the Affordable Care Act",
    align = c("c"),
    escape = FALSE,
    linesep = "",
    format = "html"
  ) %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>% 
  column_spec(2, width = "20em")

```

This is a good opportunity to see the importance of weighting in action. So far we have talked about the importance of using weights (such as sampling weights or important weights) to make statistics representative of a particular population. But we can also think about the choice of weights as a decision about how to average estimated effects between groups. California, for example, represented just 4.5% of the states that expanded Medicaid in 2014, 5.4% of the counties, but 27.7% of the adults ages 20-64. Thus its contribution to the average outcome for the 2014 expansion group is very different when weighting by adult population.

The outcome variable is the crude adult mortality rate, $Y_{it}$, for people ages 20-64 (measured per 100,000) by county (*i*) from 2009 to 2019 released by the Centers for Disease Control and Prevention (2024). The weighting variable is county *i*’s adult population in 2013, denoted by $W_i$, and socioeconomic covariates in year *t* are denoted by $X_{it}$. Table 1 defines the treatment group variable $G_i$ equal to the year in which county *i*’s state expanded Medicaid. The final sample contains 2,604 counties in states with complete data on mortality rates from 2009 to 2019 and covariates for 2013 and 2014. 

Now that we have described the setup and data, we have to think about our empirical strategy and the comparisons we want to make. Which treatment groups in Table 1 should be compared with each other, over what time horizons? What conditions must hold in order for those comparisons to identify causal effects? How can we check if these conditions are empirically plausible and assess robustness to violations of these conditions?

<br>


# $2 \times 2$ DID designs

In Quant II, you learned the DID method from the Card and Krueger (1995) minimum wage study. This is an example of the the $2 \times 2$ DID design, with two time periods (before and after the treatment) and two groups that are untreated in the first period but vary in their treatment exposure during the second period. To fix ideas, in this Medicaid example we will focus on comparing the 2014 expansion group of counties to the non-expansion group in 2013 and in 2014. We can think of this $2 \times 2$ design as the building block of more complex designs. With this $2 \times 2$ design, we can define a causal parameter for our treatment variable, the average effect of the treatment on the treated for 2014 expansion states, denoted $ATT(2014)$, along with the identifying assumption necessary for a particular estimation approach to generate an unbiased estimate of this causal parameter. Note that the *weighted* $ATT(2014)$ answers the question of “what was the average causal effect of Medicaid expansion on 2014 mortality rates *among adults in counties in states that expanded Medicaid in 2014*?” For comparison, the unweighted $ATT(2014)$ answers the question, “what was the average causal effect of Medicaid expansion on 2014 mortality rates *among the 2014 expansion state counties*?”

The identifying assumption here is the known as the *parallel trends assumption*: in the absence of treatment, the change in average outcomes for the expansion counties (the counterfactual) would be the same as for the non-expansion counties.

The estimation approach for this causal parameter under the parallel trends assumption is the familiar difference in the before-after differences for the 2014 expansion group and the non-expansion group:

\begin{align}
  \hat{ATT}(2014) = (\bar{Y}_{D_i = 1, t = 2014} -  \bar{Y}_{D_i = 1, t = 2013}) -
(\bar{Y}_{D_i = 0, t = 2014} -  \bar{Y}_{D_i = 0, t = 2013}) (\#eq:label)
\end{align}

\noindent where $D_i$ indicates whether a county is located in a 2014 expansion state, and weights are suppressed for notational ease. Parallel trends requires the counterfactual change in outcomes for the treatment group (expansion counties), to be equal to the observed change in outcomes for the comparison group (non-expansion counties) given by the latter difference in the above expression.

Use the following code to create a table showing the DID estimates, both unweighted and population-weighted:

```{r}
# set the covariates that we're going to use.
covs <- c("perc_female","perc_white", "perc_hispanic", "unemp_rate", "poverty_rate", "median_income")

## Clean data and add in covariates
mydata <- mydata %>% 
  # make variables
  mutate(perc_white = population_20_64_white / population_20_64 * 100,
         perc_hispanic = population_20_64_hispanic / population_20_64 * 100,
         perc_female = population_20_64_female/ population_20_64 * 100,
         unemp_rate = unemp_rate * 100,
         median_income = median_income / 1000) %>% 
  # keep just subset of variables that we will use later
  select(state, county, county_code, year, population_20_64, adopt, yaca,
         starts_with("perc_"), crude_rate_20_64, all_of(covs))

# keep only counties with full observations for outcome and covariates in 2013 and 2014
mydata <- mydata %>%
  # allow the aca expansion variable to be missing
  drop_na(!yaca) %>%
  group_by(county_code) %>% 
  # need full covariates for 2013 and 2014
  filter(length(which(year == 2013 | year == 2014)) == 2) %>% 
  ungroup()

# finally, keep only counties with full mortality data for 2009 to 2019
mydata <- mydata %>% 
  group_by(county_code) %>% 
  drop_na(crude_rate_20_64) %>% 
  filter(n() == 11)

# make a smaller dataset with only years 2013 and 2014
short_data <- mydata %>% 
  # make a binary variable that identifies ACA expansion and post-years
  mutate(Treat = if_else(yaca == 2014 & !is.na(yaca), 1, 0),
         Post = if_else(year == 2014, 1, 0)) %>% 
  # filter years for just 2013 and 2014
  filter(year %in% c(2013, 2014)) %>% 
  # make a variable with population weight in 2013
  group_by(county_code) %>% 
  mutate(set_wt = population_20_64[which(year == 2013)]) %>% 
  ungroup()

# get group means for Table 2
# unweighted expansion pre-reform
T_pre <- short_data %>% 
  filter(Treat == 1 & year == 2013) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# unweighted non-expansion pre-form 
C_pre <- short_data %>% 
  filter(Treat == 0 & year == 2013) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# unweighted expansion post-reform
T_Post <- short_data %>% 
  filter(Treat == 1 & year == 2014) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# unweighted non-expansion post-reform
C_Post <- short_data %>% 
  filter(Treat == 0 & year == 2014) %>% 
  summarize(mean = mean(crude_rate_20_64)) %>% 
  pull(mean)

# get the same group means but this time weighted
T_pre_weight <- short_data %>% 
  filter(Treat == 1 & year == 2013) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

C_pre_weight <- short_data %>% 
  filter(Treat == 0 & year == 2013) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

T_Post_weight <- short_data %>% 
  filter(Treat == 1 & year == 2014) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

C_Post_weight <- short_data %>% 
  filter(Treat == 0 & year == 2014) %>% 
  summarize(mean = weighted.mean(crude_rate_20_64, w = set_wt)) %>% 
  pull(mean)

# function to do a reasonable rounding
g_round <- function(x, k) {format(round(x, k), nsmall = k)}

# make table and export
tribble(
  ~"", ~"Expansion", ~ "No Expansion", ~"Gap/DiD", ~"Expansion", ~"No Expansion", ~"Gap/DiD",
  "2013", as.character(g_round(T_pre, 1)), as.character(g_round(C_pre, 1)), as.character(g_round(T_pre - C_pre, 1)), 
  as.character(g_round(T_pre_weight, 1)), as.character(g_round(C_pre_weight, 1)), as.character(g_round(T_pre_weight - C_pre_weight, 1)),
  "2014", as.character(g_round(T_Post, 1)), as.character(g_round(C_Post, 1)), as.character(g_round(T_Post - C_Post, 1)), 
  as.character(g_round(T_Post_weight, 1)), as.character(g_round(C_Post_weight, 1)), as.character(g_round(T_Post_weight - C_Post_weight, 1)),
  "Trend/DiD", as.character(g_round(T_Post - T_pre, 1)), as.character(g_round(C_Post-C_pre, 1)), 
  as.character(g_round((T_Post - T_pre) - (C_Post-C_pre), 1)),
  as.character(g_round(T_Post_weight - T_pre_weight, 1)), as.character(g_round(C_Post_weight-C_pre_weight, 1)), 
  as.character(g_round((T_Post_weight - T_pre_weight) - (C_Post_weight-C_pre_weight), 1))
) %>% 
  # format table
  kable(align = 'c',
        booktabs = T, 
        escape = F,
        caption = "Simple 2 X 2 DiD", 
        format = "html") %>% 
  kable_styling() %>% 
  row_spec(3, color = "BrickRed") %>% 
  column_spec(c(1:3, 5:6), color = "black") %>% 
  row_spec(3, italic = TRUE) %>% 
  column_spec(c(4, 7), italic = TRUE) %>% 
  add_header_above(c(" " = 1, "Unweighted Averages" = 3, "Weighted Averages" = 3))
```


## QUESTION: According to Table 2.1, what was the average treatment effect of Medicaid expansion on mortality in 2014 among *counties* that are part of an expansion state? (hint: look at the results of unweighted averages). By contrast, what was the average treatment effect on mortality in 2014 for the average *adult* in expansion states? Be explicit about the units.

::: {style="color:deepskyblue;"}
**Solution:**

Without weighting, average county-level mortality rates rose by 8.8 deaths per 100,000 in expansion states and 9.5 deaths in non-expansion states, so the DID estimate of $ATT(2014)$ is 0.7 deaths per 100,000 (after rounding). In other words, the average treatment effect of Medicaid expansion on mortality in 2014 among counties that are part of an expansion state was a decrease of 0.7 deaths per 100,000. In contrast, the population-weighted DID estimate implies that Medicaid expansion caused a reduction of 3.1 deaths per 100,000 for the average adult in expansion states—provided that the identifying parallel trends assumption holds.
:::

<br>

The same result can be obtained as the (weighted) least squares estimate of $\beta_{2x2}$ from the following PRF:

$$
  Y_{it} = \beta_0 + \beta_1 1\{D_i = 1\} + \beta_2 1\{t = 2014\} + \beta^{2x2} (1\{D_i = 1\} \times  1\{t = 2014\}) + \epsilon_{it}
$$

## QUESTION: To see why $\hat{\beta}^{2x2} = \hat{ATT}(2014)$, write out the SRF for each of the four groups defined by the treatment ($D_i$) and year. Following the first mean below, fill in the rest to show that the following holds true:

\begin{align}
\hat{ATT}(2014) = (\bar{Y}_{D_i = 1, t = 2014} -  \bar{Y}_{D_i = 1, t = 2013}) -
(\bar{Y}_{D_i = 0, t = 2014} -  \bar{Y}_{D_i = 0, t = 2013}) = \hat{\beta}^{2x2}
\end{align}

<br>

- The sample average outcome in the post-period for the treatment group is $\bar{Y}_{D_i = 1, t = 2014} = \hat{\beta}_0 + \hat{\beta}_1 + \hat{\beta}_2 + \hat{\beta}^{2x2}$

::: {style="color:deepskyblue;"}
- The sample average outcome in the pre-period for the treatment group is $\bar{Y}_{D_i = 1, t = 2013} = \hat{\beta}_0 + \hat{\beta}_1$
- The sample average outcome in the post-period for the comparison group is $\bar{Y}_{D_i = 0, t = 2014} = \hat{\beta}_0 + \hat{\beta}_2$
- The sample average outcome in the pre-period for the comparison group is $\bar{Y}_{D_i = 0, t = 2013} = \hat{\beta}_0$

With balanced panel data, the estimate of $\beta^{2x2}$ from the above PRF is numerically identical to the same coefficient from a regression that instead contains fixed effects for each county and year, also known as the Two-Way Fixed Effects (TWFE) model.

:::


## QUESTION: What are the advantages to regression-based DID estimation as opposed to relying on differences in sample averages (as we did in Question 2.2)? 

::: {style="color:deepskyblue;"}
**Solution:**

One important advantage of the regression-based estimation approach is standard error estimation. For the remainder of this lesson standard errors will be clustered at the county level. A more complete discussion of uncertainty and standard error estimation can be found in the paper.
:::

<br>


# Incorporating covariates

There are a number of reasons to consider covariates $X_{it}$ in a DID design:

1. Checking for balance in variables that could influence the counterfactual outcome; if any relevant covariates are imbalanced, that would invalidate the parallel trends assumption;
2. Controlling for these relevant covariates; or
3. Estimating heterogeneous treatment effects.

For example, in the absence of Medicaid expansion, mortality rates likely would have evolved differently in poorer and richer counties, as they did before 2014 (Currie and Schwandt, 2016). Therefore, parallel trends may fail if poverty rates differ between expansion and non-expansion counties. If they do, then one may want to “control for” poverty rates when estimating ATT parameters.


## QUESTION: Show only the mean values of each covariate in 2013 by year of adoption (by first filtering on `year == 2013`). Round to 1 decimal point. Note the average poverty rate in 2013 for counties that expanded in 2014 and non-expansion counties.

::: {style="color:deepskyblue;"}
**Solution:**

```{r}
mydata %>% 
  filter(year == 2013) %>% 
  group_by(adopt) %>% 
  summarise(across(all_of(covs), mean, na.rm = TRUE)) %>% 
  mutate(across(where(is.numeric), ~round(., 2)))
```

We observe that non-expansion counties had an average poverty rate of just below 20 percent in 2013, compared an average poverty rate of 16.4 percent in counties that expanded in 2014.
:::

<br> 


The parallel trends assumption cannot be formally tested, as the counterfactual is never observed. Thus any "test" of parallel trends is an indirect one that relies on other observed variables that could influence outcomes in the absence of the treatment ("untreated potential outcome trends"). In this case, demographic and economic conditions were strongly correlated with both mortality levels and mortality trends over time. Parallel trends would fail to hold if, for example, if county poverty rates were correlated with mortality levels and trends, and with expansion status. One way to evaluate parallel trends is to check for balance in observable determinants of potential outcomes such as poverty rates. Baker et al (2025) suggest using a normalized measure of imbalance that is comparable across variables with different underlying units and distributions: the normalized difference in means between the treatment and comparison groups (Imbens and Rubin, 2015, Chapter 14):

$$\text{Normalized difference} = \dfrac{\bar{X}_T - \bar{X}_C}{\sqrt{(S^2_T + S^2_C)/2}}$$
where $S^2_T$ and $S^2_C$ are the sample weighted variances of the covariates for the treatment and comparison group. Baker et al (2025) recommend following the general rule of thumb that values of the normalized difference in excess of 0.25 in absolute value indicate a potentially problematic imbalance between the two groups (Imbens and Rubin, 2015, page 277).


## QUESTION: Using the formula above, calculate the normalized difference in unweighted median income between treatment and comparison groups for the year 2013 from `short_data`. Round to 2 decimal points. Is it above the 0.25 rule of thumb and suggestive of problematic imbalance?

::: {style="color:deepskyblue;"}
**Solution:**

```{r}

unweighted_stats <- short_data %>% 
  group_by(year, Treat) %>% 
  summarize(mean_UR = mean(median_income),
            var_UR = var(median_income))

xt <- unweighted_stats[2,3]
xc <- unweighted_stats[1,3]
st <- unweighted_stats[2,4]
sc <- unweighted_stats[1,4]

normalized_difference <- round((xt - xc)/(sqrt((st + sc)/2)), 2)

normalized_difference[1,1]
```
Yes, the normalized difference in unweighted median income of `r normalized_difference[1,1]` is greater than 0.25, suggesting of problematic imbalance.
:::

<br>


## QUESTION: Table 3.1 below reports the means and normalized difference in means for various covariates. The top panel reports the averages and standardized differences of each variable, measured in 2013, by adoption status. The bottom panel report the average and standardized differences of the county-level long differences between 2014 and 2013 of each variable. Looking at the top panel, which baseline measures appear to be meaningfully imbalanced? Looking at the bottom panel, which baseline imbalances changed from 2013 to 2014? What does this say about the parallel trends assumption? 
```{r}
# Now, make the covariate balance table (Table 4).
# unweighted - pre
unweighted <- short_data %>% 
  filter(year == 2013) %>% 
  select(Treat, all_of(covs)) %>% 
  group_by(Treat) %>% 
  # get mean and standard deviation
  summarize_all(list(mean, var)) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_fn1 - group0_fn1)/sqrt((group1_fn2 + group0_fn2)/2)) %>% 
  select(variable, group0_fn1, group1_fn1, norm_diff)

# make a weighted variance function
wtd.var <- function (x, weights = NULL, normwt = FALSE, na.rm = TRUE, 
                     method = c("unbiased", "ML")) 
{
  method <- match.arg(method)
  if (!length(weights)) {
    if (na.rm) 
      x <- x[!is.na(x)]
    return(var(x))
  }
  if (na.rm) {
    s <- !is.na(x + weights)
    x <- x[s]
    weights <- weights[s]
  }
  if (normwt) 
    weights <- weights * length(x)/sum(weights)
  if (normwt || method == "ML") 
    return(as.numeric(stats::cov.wt(cbind(x), weights, method = method)$cov))
  sw <- sum(weights)
  if (sw <= 1) 
    warning("only one effective observation; variance estimate undefined")
  xbar <- sum(weights * x)/sw
  sum(weights * ((x - xbar)^2))/(sw - 1)
}

# weighted  - pre
weighted <- short_data %>% 
  filter(year == 2013) %>% 
  select(Treat, all_of(covs), set_wt) %>% 
  group_by(Treat) %>% 
  # get mean and standard deviation
  summarize(across(all_of(covs), 
                   list(
                     ~weighted.mean(x = ., w = set_wt),
                     ~wtd.var(x = ., weights = set_wt, normwt = TRUE)))) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_1 - group0_1)/sqrt((group1_2 + group0_2)/2)) %>% 
  select(variable, group0_1, group1_1, norm_diff)

# make the top panel (weighted and unweighted Pre)
top_panel <- bind_cols(unweighted, weighted %>% select(-variable))

# make the bottom panel, which is the same thing but with the difference in X between 2014 and 2013.
# unweighted
unweighted <- short_data %>% 
  select(county_code, year, Treat, all_of(covs)) %>% 
  arrange(county_code, year) %>% 
  group_by(county_code, Treat) %>% 
  summarize(
    across(all_of(covs), function(x) x[2] - x[1])
  ) %>% 
  ungroup() %>% 
  select(-county_code) %>% 
  group_by(Treat) %>% 
  # get mean and standard deviation
  summarize_all(list(mean, var)) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_fn1 - group0_fn1)/sqrt((group1_fn2 + group0_fn2)/2)) %>% 
  select(variable, group0_fn1, group1_fn1, norm_diff)

# weighted
weighted <- short_data %>% 
  select(county_code, year, Treat, all_of(covs)) %>% 
  arrange(county_code, year) %>% 
  group_by(county_code, Treat) %>% 
  summarize(
    across(all_of(covs), function(x) x[2] - x[1])
  ) %>% 
  ungroup() %>% 
  left_join(short_data %>% filter(year == 2013) %>% select(county_code, set_wt),
            join_by(county_code)) %>% 
  select(-county_code) %>% 
  group_by(Treat) %>%
  # get mean and standard deviation
  summarize(across(all_of(covs), 
                   list(
                     ~weighted.mean(x = ., w = set_wt),
                     ~wtd.var(x = ., weights = set_wt, normwt = TRUE)))) %>%
  # pivot the data longer
  pivot_longer(cols = !Treat, 
               names_to = "variable", 
               values_to = "value") %>% 
  # now make separate columns for treated and untreated
  pivot_wider(names_from = "Treat", 
              values_from = "value",
              names_prefix = "group") %>% 
  # separate mean and standard deviations
  extract(variable, into = c("variable", "fx"), "(.*)_(.*)") %>% 
  pivot_wider(id_cols = variable,
              names_from = fx,
              values_from = c(group0, group1)) %>% 
  # make normalized difference
  mutate(norm_diff = (group1_1 - group0_1)/sqrt((group1_2 + group0_2)/2)) %>% 
  select(variable, group0_1, group1_1, norm_diff)

# make the bottom panel
bottom_panel <- bind_cols(unweighted, weighted %>% select(-variable))

# bind the two panels
table <- bind_rows(top_panel, bottom_panel) %>% 
  # reformat all columns to two digits
  mutate(across(-variable, \(x) scales::comma(x, accuracy = 0.01)))

# format Table 4 and export
table %>% 
  # change column names
  mutate(variable = 
           case_match(variable,
                      "perc_female" ~ "% Female",
                      "perc_white" ~ "% White",
                      "perc_hispanic" ~ "% Hispanic",
                      "unemp_rate" ~ "Unemployment Rate",
                      "poverty_rate" ~ "Poverty Rate",
                      "median_income" ~ "Median Income")) %>% 
  # format latex table
  kable(col.names = c("Variable", "Non-Adopt", "Adopt", "Norm. Diff.", "Non-Adopt", "Adopt", "Norm. Diff."),
        align = 'lcccccc',
        escape = T,
        booktabs = T,
        caption = "Covariate Balance Statistics",
        linesep = "",
        format = "html") %>% 
  kable_styling(latex_options = c("scale_down", "hold_position")) %>% 
  pack_rows("2013 Covariate Levels", 1, 6) %>% 
  pack_rows("2014 - 2013 Covariate Differences", 7, 12) %>% 
  add_header_above(c(" " = 1, "Unweighted" = 3, "Weighted" = 3))
```

::: {style="color:deepskyblue;"}
**Solution**

From the top panel, several baseline measures appear to be meaningfully imbalanced: expansion counties in 2013 were whiter and had higher unemployment rates yet lower poverty and higher median income than non-expansion counties. These imbalances will lead to violations of parallel trends to the extent that counties with different racial composition or income distributions would have had different mortality changes even without Medicaid expansion.

From the bottom panel, many of the baseline imbalances change from 2013 to 2014. Unemployment, for example, was higher in expansion states in 2013 but fell faster. To the extent that these changes are important determinants of potential outcomes, this may suggest that parallel trends is violated. An alternative possibility is that these changes were actually *caused* by expansion, rather than the other way around. Whether a particular factor should be considered as a covariate or a mechanism can be a tricky, context-specific matter.
:::

## The conditional parallel trends assumption {-}

Given the observed imbalance in covariates that could invalidate parallel trends, we can formulate a new conditional parallel trends assumption which states that parallel trends holds within each covariate specific stratum—combinations of discrete covariate values—rather than over the whole population. Given the results from Table 4, this amounts to assuming parallel trends in potential outcomes between expansion and non-expansion counties that in 2013 had the same female, white, and Hispanic shares, as well as the same unemployment and poverty rates and median income. This assumption does not restrict how potential outcomes changes in different covariate strata, it simply requires that covariate-specific trends are common between expansion and non-expansion counties.

We also need to observe both treated and untreated units for all covariate values (a "common support" assumption).

You can work through the math in the original paper, but the result is that under conditional parallel trends, common support, and a no anticipation (of the treatment) assumption, $ATT(2014)$ is equal to the path of outcomes experienced by the treated group minus the average path of outcomes in the comparison group for each value of the covariates, averaged over the treated group’s distribution of covariates.


## DID estimation with covariates: TWFE {-}

The most common approach to estimating $ATT(2014)$ under conditional parallel trends is to simply add the covariates to the basic TWFE model:

\begin{align}
  Y_{it} = \theta_t + \eta_i + \beta_{treat}D_{it} + X^{'}_{it} \beta_{covs} + \epsilon_{it} \tag{2.1}
\end{align}

```{r}
reg_data_2013 <- short_data %>% 
  # make long diff in y
  group_by(county_code) %>% 
  summarize(long_y = crude_rate_20_64[which(year == 2014)] - crude_rate_20_64[which(year == 2013)]) %>% 
  # merge in 2013 covariates
  left_join(short_data %>% filter(year == 2013) %>% select(county_code, state, Treat, set_wt, all_of(covs)), 
            by = "county_code")

# this dataset has the change in X between 2014 and 2013 as controls
reg_data_change <- short_data %>% 
  # make long diff in y
  group_by(county_code) %>% 
  summarize(long_y = crude_rate_20_64[which(year == 2014)] - crude_rate_20_64[which(year == 2013)]) %>% 
  # merge in change in covariate values
  left_join(short_data %>% 
              group_by(county_code) %>% 
              mutate(set_wt = set_wt[which(year == 2013)]) %>% 
              group_by(county_code, state, Treat, set_wt) %>% 
              summarize(
                across(all_of(covs), function(x) x[which(year == 2014)] - x[which(year == 2013)])
              ), by = "county_code")
# run the six models
# first unweighted - long diff no covs, 2013 covs, then change in covs
mod1 <- feols(long_y ~ Treat, data = reg_data_2013, cluster = ~county_code)
mod2 <- feols(long_y ~ Treat + .[covs], data = reg_data_2013, cluster = ~county_code)
mod3 <- feols(long_y ~ Treat + .[covs], data = reg_data_change, cluster = ~county_code)

# same thing but weighted
mod4 <- feols(long_y ~ Treat, data = reg_data_2013, weights = ~set_wt, cluster = ~county_code)
mod5 <- feols(long_y ~ Treat + .[covs], data = reg_data_2013, weights = ~set_wt, cluster = ~county_code)
mod6 <- feols(long_y ~ Treat + .[covs], data = reg_data_change, weights = ~set_wt, cluster = ~county_code)
```

The population-weighted regression estimate of $\beta_{treat}$ falls in magnitude from `r round(mod3$coefficients[2],2)` to `r round(mod6$coefficients[2],2)` when covariates are added, though the estimates are not longer statistically significant.

However, a TWFE specification can only account for differential trends related to baseline covariate levels if pre-treatment covariate values are interacted with a post-treatment dummy:

\begin{align}
  Y_{it} = \theta_t + \eta_i + \beta_{treat,2}D_{it} + (1\{t = 2014\} X^{'}_{i,t=2013}) \beta_{covs,2} + \epsilon_{it}  \tag{2.2}
\end{align}

Note the possibility of that of "bad controls" , whereby one or more of the covariates is affected by the treatment. If Medicaid expansion caused a reduction in poverty rates, for example, then including 2014 poverty rates or the 2013-2014 change in poverty rates as a covariate would introduce bias when estimating $ATT(2014)$ with TWFE. Hence the term $1\{t = 2014\} X^{'}_{i,t=2013}$ interacts *pre-treatment* covariates with a post-treatment dummy variable. However, weighting can introduce further bias (Sloczynski, 2022; Caetano and Callaway, 2024). Fortunately, there are other approaches to estimating $ATT(2014)$ under conditional parallel trends that do not rely on TWFE.


## DID estimators with covariates that target the $ATT(2014)$ {-}

One strategy is known as the regression-adjustment (RA) or outcome regression approach to DID. With this approach, *untreated outcome trends* as a function of the covariates $X_i$ are obtained by regressing the change in outcome on covariates for *untreated* units. The fitted model (SRF) is then used to generated predicted values of the change in outcomes for *all* units (treated as well as untreated). This is one approach to mimicking the counterfactual change in outcomes for treated units.

Another strategy is known as the inverse probability weighted (IPW) DID procedure. Following from the balance checks from Table 4, if imbalance in covariates is the source of parallel trends violations, then adjusting the comparison group to be balanced on covariates can address that bias. The adjustment takes the form of re-weighting the observed changes in adult mortality rates for non-expansion counties to ensure that the expansion and non-expansion counties are similar on covariates, thus addressing the compositional source of bias. IPW relies on estimating the probability of belonging to the treatment group conditional on covariates, known as the “propensity score.”

In practice, there are major advantages to combining both RA and IPW approaches in a way that leads to estimators for $ATT(2014)$ that are more robust against model misspecification; this is called the doubly robust (DR) approach (Sant’Anna and Zhao, 2020). Between RA, IPW, and DR DID estimators, Baker et al (2025) recommend that practitioners favor DR due to the added protection from model misspecification.

The preferred DR estimator for DID is commonly referred to as the Callaway and Sant'Anna (CS) estimator, which can be estimated using the `drdid` package in R. 


## QUESTION: Obtain the unweighted and weighted $ATT(2014)$ using the doubly robust Callaway and Sant'Anna (2021) estimator (CS). To do this, first run the code chunk below to prepare the `data_cs` data frame. Second, in the next code chunk, fill in the appropriate `est_method` argument in the two `att_gt()` function calls that allow for doubly robust esimation. Third, remove the code chunk option `{eval = FALSE}` in order to run the estimation code.

## Hint: use the help function `?att_gt` to help you determine the appropriate syntax for the doubly robust `est_method`. {-}

::: {style="color:deepskyblue;"}
**Solution:**
```{r}
# finally, get the same estimates using Sant'Anna and Zhao (2020) and Callaway and Sant'Anna (2021) using the 
# three adjustment methods
# First you need to reformat the group variable (untreated = 0) and the unit ID variable needs to be numeric
data_cs <- short_data %>% 
  mutate(treat_year = if_else(yaca == 2014 & !is.na(yaca), 2014, 0),
         county_code = as.numeric(county_code))
```

```{r}
# unweighted ATT using doubly robust (DR) method
att_gt(yname = "crude_rate_20_64",
       tname = "year",
       idname = "county_code",
       gname = "treat_year",
       xformla =  as.formula(paste("~", paste(covs, collapse = "+"))),
       data = data_cs,
       panel = TRUE,
       control_group = "nevertreated",
       bstrap = TRUE,
       cband = TRUE,
       est_method = "dr",
       base_period = "universal",
       biters = 25000)

# weighted ATT using doubly robust (DR) method
att_gt(
    yname = "crude_rate_20_64",
    tname = "year",
    idname = "county_code",
    gname = "treat_year",
    xformla =  as.formula(paste("~", paste(covs, collapse = "+"))),
    data = data_cs,
    panel = TRUE,
    control_group = "nevertreated",
    bstrap = TRUE,
    cband = TRUE,
    est_method = "dr",
    weightsname = "set_wt",
    base_period = "universal",
    biters = 25000
  )
```
:::

<br>

The results from the three approaches (RA, IPW, and DR) are summarized in the following table, which reports unweighted and population-weighted 2x2 DID estimates comparing 978 counties in states that expand Medicaid in 2014 to 1,222 counties in states that did not expand Medicaid by 2019, adjusting for 2013 covariate values.


```{r}
# create a function to run the CS estimator allowing the estimation method to differ
run_cs <- function(method, wt) {
  
  # estimate the att_gt
  atts <- att_gt(
    yname = "crude_rate_20_64",
    tname = "year",
    idname = "county_code",
    gname = "treat_year",
    xformla =  as.formula(paste("~", paste(covs, collapse = "+"))),
    data = data_cs,
    panel = TRUE,
    control_group = "nevertreated",
    bstrap = TRUE,
    cband = TRUE,
    est_method = method,
    weightsname = wt,
    # faster_mode = TRUE,
    base_period = "universal",
    biters = 25000
  )
  
  # aggregate estimates 
  aggte(atts, na.rm = TRUE, biters = 25000) %>% 
    broom::tidy() %>% 
    filter(group == 2014) %>% 
    mutate(type = method)
  
}

# run it three ways - regression adjustment, IPW, and doubly robust
# these are unweighted
ests <- map_dfr(c("reg", "ipw", "dr"), run_cs, wt = NULL)

# make a table - Panel A is with unweighted models
tablea <- ests %>% 
  # reformat the data
  select(type, estimate, std.error) %>% 
  # format the estimate and std error
  mutate(estimate = scales::number(estimate, accuracy = 0.01),
         std.error = paste0("(", scales::number(std.error, accuracy = 0.01), ")")) %>% 
  # reshape the data
  pivot_longer(cols = -type,
               names_to = "statistic",
               values_to = "value") %>% 
  pivot_wider(id_cols = "statistic",
              names_from = "type",
              values_from = "value") %>% 
  mutate(statistic = if_else(statistic == "estimate", "Medicaid Expansion", " "))

# make weighted results for the three CS models.
ests_w <-  map_dfr(c("reg", "ipw", "dr"), run_cs, wt = "set_wt")

# Panel B is the same thing but with weighted results
tableb <- ests_w %>% 
  select(type, estimate, std.error) %>% 
  mutate(estimate = scales::number(estimate, accuracy = 0.01),
         std.error = paste0("(", scales::number(std.error, accuracy = 0.01), ")")) %>% 
  pivot_longer(cols = -type,
               names_to = "statistic",
               values_to = "value") %>% 
  pivot_wider(id_cols = "statistic",
              names_from = "type",
              values_from = "value") %>% 
  select(reg, ipw, dr)

# Combine Panels A and B together and format the table
bind_cols(tablea, tableb) %>% 
  kable(col.names = c(" ", "Regression", "IPW", "Doubly Robust",
                      "Regression", "IPW", "Doubly Robust"),
        align = 'lcccccc',
        escape = F,
        booktabs = T,
        caption = "Callaway and Sant'Anna (2021) DiD",
        linesep = "",
        format = "html") %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  add_header_above(c(" " = 1, "Unweighted" = 3, "Weighted" = 3))
```

## QUESTION: Focusing on the weighted estimates in the table above, how do results change across the three approaches? 

::: {style="color:deepskyblue;"}
**Solution:**

The doubly robust estimate is larger in magnitude than bot the RA and IPW estimates.
:::

<br>


# $2 \times T$ DID designs with multiple time periods (unstaggered)

The previous section developed identification strategies and estimation approaches for the $2 \times 2$ DID design—the building blocks of DID with more than two time periods. With $T > 2$ time periods there is an *ATT* in each treatment period that traces out treatment effect dynamics, as well as treated-untreated comparisons prior to the treatment that can inform the plausibility of parallel trends. 

Here the term "event study" refers to the practice of estimating effects across multiple time periods before and after the treatment. Now we expand the analysis of the 2014 Medicaid expansion group to include data from 2009 to 2019 (but still only looking at counties that expanded in 2014 or not at all before 2019). We refer to the first treatment period as *g*, such that the base period just prior to the treatment will be *t = g - 1*.


## QUESTION: First, run the code below to create `grouped_data`, which includes the weighted average of mortality rates by year by expansion group. Then, plot time series of the population-weighted morality rates for the 2014 expansion counties (treated) and post-2019 expansion counties (untreated comparison counties). Using `geom_vline()`, add a vertical line at 2014; this naturally divides the x-axis into two windows: post-treatment (2014-2019) and pre-treatment (2009-2013). 

## HINT: inspect `grouped_data` before plotting. {-}

```{r}
# load cleaned data
mydata <- read_csv("county_mortality_data.csv") %>% 
  # make state the abbreviation
  mutate(state = str_sub(county, nchar(county) - 1, nchar(county))) %>% 
  # drop DC and pre-2014 adoption states
  filter(!(state %in% c("DC", "DE", "MA", "NY", "VT"))) %>% 
  # drop states that adopt between 2014 and 2019
  filter(yaca == 2014 | is.na(yaca) | yaca > 2019)

# set the covariates that we're going to use.
covs <- c("perc_female","perc_white", "perc_hispanic", 
          "unemp_rate", "poverty_rate", "median_income")

## Clean data and add in covariates
mydata <- mydata %>% 
  # make variables
  mutate(perc_white = population_20_64_white / population_20_64 * 100,
         perc_hispanic = population_20_64_hispanic / population_20_64 * 100,
         perc_female = population_20_64_female/ population_20_64 * 100,
         unemp_rate = unemp_rate * 100,
         median_income = median_income / 1000) %>% 
  # keep just subset of variables that we will use later
  select(state, county, county_code, year, population_20_64, yaca,
         starts_with("perc_"), crude_rate_20_64, all_of(covs))

# keep only counties with full observations for outcome and covariates in 2013 and 2014
mydata <- mydata %>%
  # allow the aca expansion variable to be missing
  drop_na(!yaca) %>%
  group_by(county_code) %>% 
  # need full covariates for 2013 and 2014
  filter(length(which(year == 2013 | year == 2014)) == 2) %>% 
  ungroup()

# finally, keep only counties with full mortality data for 2009 to 2019
mydata <- mydata %>% 
  group_by(county_code) %>% 
  drop_na(crude_rate_20_64) %>% 
  filter(n() == 11)

# make ACA and post variables, as well as the weighting variable which is the relevant county 
# population in 2013
mydata <- mydata %>% 
  mutate(Treat = if_else(yaca == 2014 & !is.na(yaca), 1, 0),
         Post = if_else(year >= 2014, 1, 0)) %>% 
  group_by(county_code) %>% 
  # make a variable with population weight in 2013
  mutate(set_wt = population_20_64[which(year == 2013)]) %>% 
  ungroup()

grouped_data <- mydata %>% 
  mutate(expand = if_else(Treat == 1, "Expansion Counties", "Non-Expansion Counties")) %>% 
  # get the weighted averages by expansion type and year
  group_by(expand, year) %>% 
  summarize(mortality = weighted.mean(crude_rate_20_64, set_wt))  
```


::: {style="color:deepskyblue;"}
**Solution:**

```{r}
# make time series plot
grouped_data %>% 
  ggplot(aes(x = year, y = mortality, group = expand, color = expand)) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  geom_vline(xintercept = 2014, linetype = "dashed") + 
  scale_color_brewer(palette = 'Set1') + 
  scale_x_continuous(breaks = 2009:2019) + 
  labs(x = "", y = "Mortality (20-64) \n Per 100,000") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```
:::

<br>

## Event study estimates in the post-treatment periods {-}

In the simple event study, the target parameters are the familiar $ATT(t)$'s from the $2 \times 2$ design, but now there are many more of them to identify, estimate, and interpret. But identifying each $ATT(t)$ requires parallel trends in every post-period, which we will refer to as the event study parallel trends assumption (Assumption PT-ES): in the absence of the treatment, the average change in outcomes for the treated group between period *t* and one period prior to the treatment (*g - 1*) is the same as for untreated group, for every post-treatment period $t$.

Learning about dynamic effects over more than two time periods thus requires stronger assumptions than the $2 \times 2$ case. In order to identify the average effect of Medicaid expansion in 2019 among expansion counties, Assumption PT-ES requires parallel trends to hold in every year from 2014 to 2019. By comparison, if we are only interested in learning about short-run effects—for example, up until 2015—then we would only require parallel trends to hold from 2014 and 2015.

If Assumption PT-ES holds, as well as a no-anticipation assumption, then every $ATT(t)$ is identified by a DID comparison between period $g-1$ and $t$, and the $ATT(t)$ can be estimated by the familiar difference in the difference of sample average outcomes (population weights are again omitted for notational ease):

\begin{align}
  \hat{ATT}(t) = (\bar{Y}_{D=1,t} -  \bar{Y}_{D=1,t=g-1}) -
(\bar{Y}_{D=0,t} -  \bar{Y}_{D=0,t=g-1}) \tag{4.1}
\end{align}

The figure below plots (population-weighted) event study estimates for the 2014 Medicaid expansion group. The $\hat{ATT}(t)$s are plotted to the right of the vertical dashed line dropped just at the reference period, 2013 ($t = g - 1$). Note that the estimation for event time 0 is the same as the $2 \times 2$ result from Table 2 (-2.6). The remaining estimates during the treatment period have the same form: comparisons of between-group (treated to untreated) changes in different post-2014 years ($t$) but always relative
to 2013 ($g-1$).


```{r}
# Make event study plots
# first add in the variables needed to run the CS model - timing group variable (treat_year)
# and county code needs to be numeric
mydata <- mydata %>% 
  mutate(treat_year = if_else(yaca == 2014 & !is.na(yaca), 2014, 0),
         county_code = as.numeric(county_code),
         time_to_treat = if_else(Treat == 1, year - treat_year, 0))

# estimate CS models

# get the individual ATT(g,t) estimates
mod <- did::att_gt(
  yname = "crude_rate_20_64",
  tname = "year",
  idname = "county_code",
  gname = "treat_year",
  xformla =  NULL,
  data = mydata,
  panel = TRUE,
  control_group = "nevertreated",
  bstrap = TRUE,
  cband = TRUE,
  est_method = "reg",
  weightsname = "set_wt",
  # faster_mode = TRUE,
  base_period = "universal",
  biters = 25000
)

# confirm you get the same with OLS (standard errors differ because of bootstrap only)
cs_out <- mod

ols_out <- feols(crude_rate_20_64 ~ i(time_to_treat, Treat, ref = -1) | county_code + year, 
      data = mydata, 
      cluster = ~county_code, 
      weights = ~set_wt)

# get the Rambachan/Roth confidence interval
# first save our aggregate event study estimates
es <- did::aggte(
  mod,
  type = "dynamic",
  min_e = -5,
  max_e = 0,
  bstrap = TRUE,
  biters = 25000
)

# source the honest_did code made to fit our estimates
source(here::here('5_honestdid.R'))

# get robust CI
robust_ci <- honest_did(es = es, type = "relative_magnitude")$robust_ci

# get the aggregate value for e = 0:5
agg <- mod %>% 
  aggte(type = "dynamic", min_e = 0, max_e = 5,
        bstrap = TRUE, biters = 25000)

# make labels for the estimate, std error, and confidence intervals
label1 <-  paste0("Estimate~(t %in% '{0, 5}')~'='~'", scales::number(agg$overall.att, accuracy = 0.01), "'")

label2 <- paste0("Std. Error = ", scales::number(agg$overall.se, 0.01), " \n",
                 "Conf. Int = [", scales::number(agg$overall.att - 1.96*agg$overall.se, 0.01), ", ", 
                 scales::number(agg$overall.att + 1.96*agg$overall.se, 0.01), "]")

# make event study plot for Figure 
mod %>% 
  # Aggregate in event time ("dynamic")
  aggte(type = "dynamic", biters = 25000) %>% 
  # get the two confidence intervals
  broom::tidy(conf.int = TRUE) %>% 
  # plot
  ggplot(aes(x = event.time, y = estimate)) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), color = "darkred") + 
  geom_linerange(aes(ymin = point.conf.low, ymax = point.conf.high)) + 
  geom_point() + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_continuous(breaks = -5:5) + 
  annotate("text", x = 3, y = 11, label = label1, parse = TRUE) + 
  annotate("text", x = 3, y = 9, label = label2) + 
  labs(x = "Event Time", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```


## QUESTION: Based on the above event study plot, are any of the estimated treatment effects for specific relative time periods statistically significant? Be specific about which relative time periods 𝑗={0,1,2,3,4,5} have estimated effects that are/aren’t significant, and explain how you can determine this from the above plot.

::: {style="color:deepskyblue;"}
**Solution:**

None of the estimates for the treatment period are statistically significant, as evidenced by the fact that the horizontal axis (y = 0) intersects the the confidence interval for every estimate of $\hat{ATT}(t)$ where $t > g - 1$.
:::

<br>


## Event study estimates in the pre-periods {-}

Multiple time periods in the pre-treatment period also allow us to perform a falsification test that informs the plausibility of the parallel trends and no anticipation assumptions, which implies every $ATT(t)$ before time period $g$ is equal to 0. We now refer to the *ATT*s between the reference period $t=g-1$ and $t=g-k$ going back *k* periods prior to the treatment as $\tau_{-k}$, to distinguish these "pre-trends" from actual treatment effects during the treatment period.

Thus the internal validity of an event study depends in large part on finding small estimates of $\tau_{-k}$. Indeed, from the event study plot we can see that none of these estimates are statistically significant. But that doesn't mean that parallel trends holds; parallel trends is inherently untestable, as we never observe the counterfactual outcome change for treated units. In this case, we fail to reject the null hypothesis that the pre-trend estimate equals zero (individually or jointly). But the pre-trend estimates also trend to be positive and larger in magnitude than many of the post-period estimates. What do we make of these pre-trends, and what does it say about the viability of the parallel trends assumption?

Baker et al (2025) caution against relying solely on an eye-test of whether pre-trends are meaningfully different from zero. A better approach begins with finding the maximum deviation from parallel trends observed in the pre-period. Next, find how large a deviation could occur in the post-period (relative to the maximum pre-period violation) before a significant *ATT* breaks down. These methods have been formalized by Rambachan and Roth (2023) and can be implemented using the `HonestDID` R package.


## Estimation and aggregating effects across time in event-studies {-}

An event study can be understood as a collection of simple $2 \times 2$ DID comparisons, one for each time period relative to treatment. Rather than estimating each of these separately, researchers can recover all period-specific treatment effects in a single step by running a two-way fixed effects (TWFE) regression with unit fixed effects, time fixed effects, and interactions between the treatment group indicator and time dummies. One period, typically the period immediately before treatment, is omitted to serve as the baseline, so each coefficient measures the treatment effect relative to that reference period. This splits the coefficients into pre- and post-treatment groups, as in the following PRF:

\begin{align}

Y_{i,t} = \theta_t + \eta_i + \sum_{k=1}^{g=-2} \beta_k^{pre} 1\{G_i = g\} \cdot 1\{t = k\} + \sum_{k=g}^{T} \beta_k^{post} 1\{G_i = g\} \cdot 1\{t = k\} + \epsilon_{it} \tag{4.2}
\end{align}

Estimates from equation (4.2) produces the same point estimates as the underlying $2 \times 2$ DID calculations from equation (4.1), along with standard errors and confidence intervals. 

An additional issue in event study inference, however, involves the fact that we are now estimating many treatment effect parameters. Thus, when we compare across event study estimates, we are conducting many hypothesis tests, and the usual normal critical values used to construct confidence intervals do not account for these. Asymptotically correct inferences about the entire event study curve require “inflating” critical values to perform a multiple hypothesis test adjustment.

A related issue arises when aggregating these period-specific effects into a single summary measure, such as the average treatment effect across post-treatment periods. While this can be computed directly by averaging the event study estimates, a common shortcut is to estimate a simpler regression with a single treatment indicator, in place of multiple event study dummies. This approach does not generally recover the same parameter, because it  averages outcomes across all pre- and post-treatment periods, rather than comparing each post-treatment period to a consistent baseline. As a result, the estimate may be biased if the parallel trends assumption does not hold uniformly across all pre-treatment periods.


## Covariates in event studies {-}

Another advantage of seeing event studies as collections of $2 \times 2$ building blocks is that all the tools for incorporating covariates immediately apply to each event study estimate. The key difference is that, instead of comparing short changes over one period, we now compare “long differences:” the change in outcomes from the baseline period (just before treatment) to each later time period. To incorporate covariates, researchers can use familiar approaches such RA, IPW, or DR. Each of these methods adjusts for differences in observable characteristics between treated and untreated units, but they do so in slightly different ways: RA models outcomes directly, IPW re-weights observations based on treatment probabilities, and DR combines both approaches. In an event study, these adjustments must be applied separately for each time period.

Note that the same concerns that arise in simpler DID settings, such as model misspecification or inappropriate controls, still apply, and can become even more complicated with multiple time periods. 



# Staggered treatment adoption ($G_{\#} \times T$)

Now that we understand $2 \times T$ event studies as a collection of $2 \times 2$ DID building blocks, we can extend the setup to consider a staggered treatment rollout. With staggered treatment timing, each treatment date defines a distinct treatment group with its own $2 \times T$ design and event study parameters. But we have to think carefully about selecting the comparison units for identifying and estimating group-specific event studies, and how to aggregate the estimates across timing groups.

Extending the framework requires a modified No Anticipation assumption which states that treatment effects are zero in all pre-treatment periods owing to units not acting on the potential knowledge of future treatment dates before they are actually exposed to the treatment.

We must also assume that a “never-treated” group always exists in our staggered DID setup. This means that if all units are eventually treated, we must drop all observations for the period corresponding to the last treatment cohort, so that units treated in this last cohort are actually never treated during the *T* periods of data that is actually used in the analysis.

We can plot weighted mortality by year for each treatment year (year of Medicaid expansion), as in Figure 5 below: 

```{r}
# set seed for reproducibility
set.seed(20240924)

# load cleaned data
mydata <- read_csv(here::here("data", "county_mortality_data.csv")) %>% 
  # make state the abbreviation
  mutate(state = str_sub(county, nchar(county) - 1, nchar(county))) %>% 
  # drop DC and pre-2014 adoption states
  filter(!(state %in% c("DC", "DE", "MA", "NY", "VT"))) 

# set the covariates that we're going to use.
covs <- c("perc_female","perc_white", "perc_hispanic", "unemp_rate", "poverty_rate", "median_income")

## Clean data and add in covariates
mydata <- mydata %>% 
  # make variables
  mutate(perc_white = population_20_64_white / population_20_64 * 100,
         perc_hispanic = population_20_64_hispanic / population_20_64 * 100,
         perc_female = population_20_64_female/ population_20_64 * 100,
         unemp_rate = unemp_rate * 100,
         median_income = median_income / 1000) %>% 
  # keep just subset of variables that we will use later
  select(state, county, county_code, year, population_20_64, yaca,
         starts_with("perc_"), crude_rate_20_64, all_of(covs))

# keep only counties with full observations for outcome and covariates in 2013 and 2014
mydata <- mydata %>%
  # allow the aca expansion variable to be missing
  drop_na(!yaca) %>%
  group_by(county_code) %>% 
  # need full covariates for 2013 and 2014
  filter(length(which(year == 2013 | year == 2014)) == 2) %>% 
  ungroup()

# finally, keep only counties with full mortality data for 2009 to 2019
mydata <- mydata %>% 
  group_by(county_code) %>% 
  drop_na(crude_rate_20_64) %>% 
  filter(n() == 11)

# make smaller dataset
mydata <- mydata %>% 
  mutate(Treat = if_else(!is.na(yaca) & yaca <= 2019, 1, 0),
         treat_year = if_else(!is.na(yaca) & yaca <= 2019, yaca, 0),
         Post = if_else(year >= 2014, 1, 0)) %>% 
  group_by(county_code) %>% 
  # make a variable with population weight in 2013
  mutate(set_wt = population_20_64[which(year == 2013)]) %>% 
  ungroup()

# make time series plot by timing group (Figure 5)
mydata %>% 
  # identify the groups
  mutate(treat_year = if_else(treat_year == 0, "Non-Expansion Counties", as.character(treat_year))) %>% 
  # get weighted average mortality by gtiming-group and year
  group_by(treat_year, year) %>% 
  summarize(mortality = weighted.mean(crude_rate_20_64, set_wt)) %>% 
  ggplot(aes(x = year, y = mortality, group = as.factor(treat_year), color = as.factor(treat_year))) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  scale_color_manual(values = c("#7C7189", "#D04E59", "#BC8E7D", "#2F3D70", "#CABEE9")) + 
  scale_x_continuous(breaks = 2009:2019) + 
  labs(x = "", y = "Mortality (20-64) \n Per 100,000") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

Up until now we have only considered a single treatment group, the 2014 expansion states. But we can see from Table 1 that as of 2019, there were four different groups of expansion states defined by whether they expanded Medicaid in 2014, 2015, 2016, or 2019. We must modify our notation to account for multiple treatment groups, treatment effects (ATT parameters), parallel assumptions, and estimators. 

In the simple $2 \times 2$ design, there is one target parameter ($ATT(2)$), while a $2 \times T$ design has *T-1* target ATT parameters, including $T-(G-1)$ in the pre-treatment period and $G-2$ in the treatment period. 

In a staggered DID design, each treatment group or cohort (defined by treatment date *g*), has its own set of *T-1* event study parameters known as group-time average treatment effects and denoted by $ATT(g,t)$. Each $ATT(g,t)$ is the average treatment effect of starting treatment at period g relative to never starting it, at time period t, among units that actually started treatment in period *g*. It is simply a set of event study parameters for treatment timing group *g*. The unobserved counterfactuals are now untreated potential outcome means for each treatment group in each period,


## Identification with staggered designs

Under No Anticipation, a set of parallel trends assumptions for $t \geq g$ identifies the causal post-treatment parameters. DID comparisons for $t \leq g-1$ represent differential pre-trends in untreated potential outcomes.

Now in a staggered setting, we have multiple treatment groups with different treatment starting dates to consider for alternative sets of comparison groups. For example, so far our Medicaid analysis has used counties in states that
did not expand Medicaid by 2019 as the comparison group. For estimating the *ATT* for the 2014 expansion group in 2015, counties in states that did not expand until 2016, 2017, or 2018 could also serve as comparison units. We have to think carefully about choosing which groups should be used to make comparisons against in order to identify $ATT(g,t)$, which in turn shapes the necessary parallel trends assumption. Staggered timing creates many potential parallel trends assumptions!

Three options are to use either the never-treated or not-yet-treated groups as the comparisons for all eventually-treated groups (Callaway and Sant’Anna, 2021), or to assume that parallel trends holds in all periods and in all groups. Each require a particular set of parallel trends assumptions for each treatment group. 

Never-treated units are perhaps the simplest comparison group to consider. But we can also use not-yet-treated groups (later adopters), i.e. groups that have not yet been treated as of period *t*. For example, when estimating $ATT(2014,2015)$, we can include the 2016 and 2019 treatment group as a part of the comparison group. The requisite parallel trend assumption (Parallel Trends based on not-yet-treated groups, or PT-GT-NYT), states that the difference in potential outcomes between any two post-treatment periods for each treatment group is equal to the difference in outcomes between the same two periods for units that have yet to be treated.


## Estimators for staggered designs without covariates

Under Parallel Trends based on not-yet-treated groups (PT-GT-NYT), one can derive an estimator, $\hat{ATT}_{nyt}(g,t)$. Also known as the CS estimator, from Callaway and Sant’Anna (2021). Sun and Abraham derived an alternative estimator, which can be implemented in R using the `sunab()` function in the `fixest` package.

Thus we can utilize the $2 \times 2$ DID framework to estimate heterogeneous $ATT(g,t)$ parameters, with the exact estimator
depending on which parallel trends assumption we think holds. Figure 6 (below) presents estimates of $ATT(g,t)$ under the PT-GT-NYT assumption. In this application, there are have four sets of counties defined by their Medicaid expansion timing, which corresponds to four sets of event studies, one for each expansion group. For the 2014, 2016, and 2019 expansion groups, the results imply Medicaid did not lead to significant changes in adult mortality rates. For the 2015 expansion group, adult mortality rates rose after expansion. Figure 6 also suggests that there may be some non-negligible pre-trends for the 2016 expansion group, though they are not statistically different from zero.

```{r}
# reclassify the county code as numeric for Callaway/Sant'Anna
mydata <- mydata %>% 
  mutate(county_code = as.numeric(county_code))

# get the ATT(g,t)'s
mod_no_x <- att_gt(
  yname = "crude_rate_20_64",
  tname = "year",
  idname = "county_code",
  gname = "treat_year",
  xformla =  NULL,
  data = mydata,
  panel = TRUE,
  control_group = "notyettreated",
  bstrap = TRUE,
  biters = 25000,
  cband = TRUE,
  weightsname = "set_wt",
  # faster_mode = TRUE,
  base_period = "universal"
)

# make plots in calendar time by timing group
mod_no_x %>% 
  # get the estimates and confidence intervals
  broom::tidy(conf.int = TRUE) %>% 
  mutate(group = as.character(group)) %>% 
  # plot the estimates and CIs
  ggplot(aes(x = time, y = estimate, color = group)) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high)) + 
  scale_x_continuous(breaks = seq(2009, 2019, by = 2),
                     labels = seq(2009, 2019, by = 2)) + 
  scale_color_manual(values = c("#7C7189", "#D04E59", "#BC8E7D", "#2F3D70")) + 
  geom_vline(aes(xintercept = as.numeric(group) - 1), linetype = "dashed", linewidth = 1) + 
  labs(x = "", y = "Treatment Effect \n Mortality Per 100,000") + 
  facet_wrap(~group) + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```


## Aggregating group-time average treatment effects

A remaining question is how to aggregate estimates of $ATT(g,t)$ from Figure 6 into a summary treatment effect measures. This aggregation would offer clear advantages in terms of improving precision, reducing the number of estimates to present and interpret, and yielding a parameter that averages over all treated units.

In a staggered design, aggregation requires a notion of time (calendar time *t* or event-time $e=t-g$), a length of time (how many periods to aggregate across), and group weights (so that larger treatment groups “matter more” than smaller ones). Given these parameters, it is straightforward to aggregate the $ATT(g,t)$ building blocks into a summary measure:

$$
ATT_{aggte} = \sum_{g,t1} \omega_{\omega,g,t} ATT(g,t),
$$
where $\omega_{\omega,g,t}$ is a generic (ω-weighted) group and time-specific (non-negative) weight that sums up to one.

Obviously positive weight can only be assigned to $ATT(g,t)$ estimates that are actually observed given the available data and treatment dates. Earlier treated groups have more post-treatment periods and thus more $ATT(g,t)$ estimates, whereas later treated groups have fewer post-treatment periods but earlier pre-treatment periods for assessing pre-trends. This means that is not possible to include every group in an aggregated event-study parameter at every event-time period $e=t-g$. To see which $ATT(g,t)$ estimate will contribute to our event study, Figure 7 stacks the event study pictures from Figure 6 and recenter them around event time instead of calendar time. In other words, Figure 7 plots $ATT(g,g+e)$ against over event time *e* for each expansion group.

```{r}
# Make a different plot that reports the ATT(g,t)s in relative time
mod_no_x %>% 
  # get the estimates and CIs
  broom::tidy() %>% 
  # define relative time for each group
  mutate(rel_time = time - group) %>% 
  # plot
  ggplot(aes(x = rel_time, y = estimate, group = as.factor(group), color = as.factor(group))) + 
  geom_point(size = 2) + geom_line(linewidth = 1) + 
  geom_vline(xintercept = -1, linetype = "dashed", linewidth = 1) + 
  scale_color_manual(values = c("#7C7189", "#D04E59", "#BC8E7D", "#2F3D70")) + 
  scale_x_continuous(breaks = -10:5) + 
  labs(x = "", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(legend.position = 'bottom',
        axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

We can take “vertical” weighted averages of each available $ATT(g,t)$ estimate for each event time period *e*. This means, for example, that to estimate an aggregate event study for for immediate treatment effects in event time 0, we can average estimates of $ATT(2014,2014)$, $ATT(2015,2015)$, $ATT(2016,2016)$ and $ATT(2019,2019)$. For event time 1, we can average estimates of $ATT(2014,2015)$, $ATT(2015,2016)$, $ATT(2016,2017)$, but not the 2019 treatment group for which there is no observable event time $e=1$ in the final dataset.

For the aggregate event study to be a representative summary of dynamic effects among treated counties, one must choose weights that are proportional to the treatment group size.

Using the CS estimator, $ATT_{cs}^e(e)$ gives the average treatment effect among the units that have been exposed to
treatment for exactly *e* periods, conditional on being observed as having participated in the treatment for that number of periods, and ever-participating in the treatment by period *T*.

Finally, one can also take a simple average of all available post-treatment event times, $ATT_{cs}^e(e), e\geq 0$, to arrive at a single, overall ATT measure.

Figure 8 reports population-weighted estimates of the event-study aggregation from event times -5 to 5, which correspond to 5 years before/after Medicaid expansion, including 95% confidence intervals. Overall, the results suggest that Medicaid expansion has no effect on adult mortality rates among counties that eventually experience a Medicaid expansion. The pre-trends are also fairly close to zero, suggesting that our parallel trends assumption may be reasonable.

```{r}
# Make a full -5 to + 5 event study plot for no covariates 
# using Callaway/Sant'Anna with not-yet-treated

# get the aggregate value for e = 0:5
agg <- mod_no_x %>% 
  aggte(type = "dynamic", min_e = 0, max_e = 5,
        bstrap = TRUE, biters = 25000)

# make labels for the estimate, std error, and confidence intervals
label1 <-  paste0("Estimate~(e %in% '{0, 5}')~'='~'", scales::number(agg$overall.att, accuracy = 0.01), "'")

label2 <- paste0("Std. Error = ", scales::number(agg$overall.se, 0.01), " \n",
                 "Conf. Int = [", scales::number(agg$overall.att - 1.96*agg$overall.se, 0.01), ", ", 
                 scales::number(agg$overall.att + 1.96*agg$overall.se, 0.01), "]")

# make G*T event study plot without covariates
mod_no_x %>% 
  # aggregate into relative time (dynamic)
  aggte(type = "dynamic") %>% 
  broom::tidy(conf.int = TRUE) %>% 
  # keep just -5 to + 5
  filter(event.time %>% between(-5, 5)) %>% 
  # plot
  ggplot(aes(x = event.time, y = estimate)) + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), color = "darkred") + 
  geom_linerange(aes(ymin = point.conf.low, ymax = point.conf.high)) + 
  geom_point() + 
  geom_vline(xintercept = -1, linetype = "dashed") + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_continuous(breaks = -5:5) + 
  annotate("text", x = 3, y = 11, label = label1, parse = TRUE) + 
  annotate("text", x = 3, y = 9, label = label2) + 
  labs(x = "Event Time", y = "Treatment Effect \n Mortality Per 100,000") + 
  theme(axis.title.y = element_text(hjust = 0.5, vjust = 0.5, angle = 360),
        strip.text = element_text(size = 14),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```


## Estimators for staggered designs with covariates

The CS estimator can be extended to incorporate covariates under a modified Assumption CPT-GT-NYT (Conditional Parallel Trends based on not-yet-treated groups), but we'll stop here for now.


# Limitations of TWFE regressions

The framework developed here builds an estimator from $2 \times 2$ building blocks, each of which targets a well-defined ATT parameter under a specific parallel trends assumption. The familiar TWFE estimator, on the other hand, comes from extending convenient estimation tools that work well in the $2 \times 2$ case. Here is a TWFE specification that estimates a summary
treatment effect parameter:

\begin{align}
  Y_{it} = \theta_t + \eta_i + \beta^{twfe}D_{it} + \epsilon_{it} \tag{5.15}
\end{align}

However, the recent DID literature has identified to major issues with $\beta^{twfe}$ in a staggered setting. The primary issue comes from the fact that TWFE implicitly uses already-treated units as a part of comparison groups. Even if PT holds for all groups and all periods, the resulting parameter can actually put negative weight on certain $ATT(g,t)$ parameters. The only way for TWFE to avoid the problem is if treatment effects do not change over time, which is a strong and often unreasonable additional assumption.

While there is an inherent simplicity to using $2 \times 2$ building blocks to develop DID identification and estimation strategies, the details can be quite complicated--especially in a staggered setting! For your projects, I recommend you begin by building out the appropriate descriptive plots. In a *staggered* setting, this means Figure 6 and 7, which stacks the different DID event study designs for different treatment groups. If you have a single treatment date for all treated groups, the path is much simpler: plot differences in mean outcomes over time as in question 4.1, wrestle with parallel trends and potential confounders, and try to build up a TWFE specification that you can defend. Some of the details you will need to wrestle with include the number of periods to consider and the choice of a weighting, if appropriate.
